Abstract Binding Trees

### 2015-11-07 paper.tex
[paper] Use element notation for map
[paper] notation adjustment in pullback square
- make the diagram well-moded (\oint O is the *result* of pulling back O
  along ...)

- use the tuple notation for elements consistently
Revert part of the previous notation change
Revamp interpretation presentation
I've arranged this to feel a bit more natural/readable & less
artificial.

\[
\begin{tikzcd}[ampersand replacement = \&, sep = large]
  {\Grothendieck{}{\Operators}}
    \arrow[rr, "{\Pair{\Pair{\Upsilon}{a}}{\vartheta}\ \mapsto\ \ArrId{\Operators\Pair{\Upsilon}{a}}}"]
    \arrow[d, "{\pi_{\Operators}}" left]
  \&
  \& {\Arr{\Sets}}
    \arrow[d, "{\pi_{cod}}"] \\
  {\SCtx\times\Discrete{\Arities}}
     \arrow[rr, "{\Operators}" below]
  \&
  \& {\Sets} \\
\end{tikzcd}
\]

@@

Next we have the interpretation of metavariable applications
\[%
  \SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash%
    \MApp{\MV{m}}{\vec{u}}{\vec{M}} : \tau}_P%
  \qquad(\Lookup{\Omega}{\MV{m}}{\MkValence{\vec{\sigma}}{\vec{\tau}}{\tau}})
\]
which proceeds by projecting the metavariable's interpretation from the
environment and intantiating it via substitution:
\[
  \begin{tikzcd}[ampersand replacement = \&, sep=huge]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
          \arrow[r, "\IMode{\Tuple{\pi_\MV{m}\,\pi_1, \phi\,\pi_2, \psi}}"]
    \& {\IMode{P_{\tau}^{\Yoneda{\vec{\sigma}\parallel\vec{\tau}}}\times S^{\{\vec{\sigma}\}} \times P^{[\vec{\tau}]}}}
          \arrow[r, "\IMode{\varsigma_{\vec{\sigma}\parallel\vec{\tau}}}"]
    \& \IMode{P_\tau}
  \end{tikzcd}
  \tag{metavar}
\]
where
\begin{align*}
  \ADefine{\phi}{
  \begin{tikzcd}[ampersand replacement = \&, sep=huge]
    {S^{\{\Upsilon\}}} \arrow[r, "\Tuple{\pi_u}_{u\in\vec{u}}"]
    \& {S^{\{\vec{\sigma}\}}}
  \end{tikzcd}
  }\\
  \ADefine{\psi}{
   \begin{tikzcd}[ampersand replacement = \&, sep=28ex]
     {\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}
     \arrow[r, "\Tuple{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash M:\tau}_P}_{(M,\tau)\in (\vec{M},\vec{\tau})}"]
     \& {P^{[\vec{\tau}]}}
   \end{tikzcd}
  }
\end{align*}

The interpretation of operator applications is most complicated. Recall that,
unlike in standard treatments of universal algebra, our operators are indexed
by symbol collections; therefore, operators must pass through suitable
renamings in order to be used in the interpretation. Let us begin by
constructing for each operator $\IsOperator{\Upsilon}{\vartheta}{a}$ the
natural transformation
$\Of{\SemBrackets{\vartheta}_P}{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P
\to\Operators_{a}}$, which shall rename the parameters of the operator using
the environment.

\[
  \Define{
    \SemBrackets{\vartheta}_P
  }{
   \begin{tikzcd}[ampersand replacement = \&, sep=large]
     {\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P} \arrow[r, "\pi_2"]
     \& {S^{\{\Upsilon\}}} \arrow[r, "\FLift{\pi_{(-)}}{}"]
     \& {\Operators_a}
   \end{tikzcd}
 }
\]

We will proceed using the initial $\SigFun{\Sigma}$-algebra $\alpha$, as
follows.
\[
   \begin{tikzcd}[ampersand replacement = \&, sep=large]
     {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}} \arrow[r, "\OMode{\beta}"]
     \& {\IMode{\SigFun{\Sigma}(P)_\tau}} \arrow[r, "\IMode{\alpha_\tau}"]
     \& {\IMode{P_\tau}}
   \end{tikzcd}
\]

\begin{align*}
  \ADefine{
    \beta
  }{
   \begin{tikzcd}[ampersand replacement = \&, sep=huge]
      {\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P} \arrow[r, "\Tuple{\SemBrackets{\vartheta}_P, \gamma}"]
      \& {
        \coprod_{\vartheta\in \Operators_{\MkArity{\vec{v}}{\tau}}}
        \prod_{\MkValence{\vec{\sigma}_i}{\vec{\tau}_i}{\tau_i}\in\vec{v}}
        P_{\tau_i}^{\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}}
      }
   \end{tikzcd}
  }\\
  \ADefine{
    \gamma
  }{
   \begin{tikzcd}[ampersand replacement = \&, sep=6em]
      {\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P} \arrow[r, "\Tuple{\lambda x. \eta_i(-,x))}_i"]
      \& {
        \prod_{\MkValence{\vec{\sigma}_i}{\vec{\tau}_i}{\tau_i}\in\vec{v}}
        P_{\tau_i}^{\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}}
      }
   \end{tikzcd}
  }\\
  \ADefine{
    \eta_i
  }{
   \begin{tikzcd}[ampersand replacement = \&, sep=7em]
      {\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P
        \times\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_u}
      } \arrow[r, "\Tuple{\pi_1\circ\pi_1,\phi_i,\psi_i}"]
      \& {\SemBrackets{\Omega\triangleright\Upsilon,\vec{\sigma}_i\parallel\Gamma,\vec{\tau}_i}_P}
      \arrow[r, "\SemBrackets{\Omega\triangleright\Upsilon,\vec{\sigma}_i\parallel\Gamma,\vec{\tau}_i\vdash M_i:\tau_i}_P"]
      \& {P_{\tau_i}}
   \end{tikzcd}
  }\\
  \ADefine{
    \phi_i
  }{
   \begin{tikzcd}[ampersand replacement = \&, sep=12em]
      {\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P
        \times\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}
      } \arrow[r, "\Tuple{\pi_u\circ\pi_1}_{u\in\Dom\Upsilon}\oplus\Tuple{\pi_u\circ\pi_1\circ\pi_2}_{u\in\Dom{\vec{\sigma}_i}}"]
      \& {S^{\{\Upsilon,\vec{\sigma}_i\}}}
   \end{tikzcd}
  }\\
  \ADefine{
    \psi_i
  }{
   \begin{tikzcd}[ampersand replacement = \&, sep=12em]
      {\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P
        \times\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}
      } \arrow[r, "\Tuple{\pi_x\circ\pi_1}_{x\in\Dom\Gamma}\oplus\Tuple{\pi_x\circ\pi_1\circ\pi_2}_{x\in\Dom{\vec{\tau}_i}}"]
      \& {V^{[\Gamma,\vec{\tau}_i]}}
   \end{tikzcd}
  }
\end{align*}

>>>>

\[
\begin{tikzcd}[ampersand replacement = \&, sep = large]
  {\OMode{\Grothendieck{}{\Operators}}}
  \arrow[rr,
    "{\IMode{\vartheta\in\Operators\Pair{\Upsilon}{a}\ \mapsto\
      \ArrId{\Operators\Pair{\Upsilon}{a}}}}"] \arrow[d, "\IMode{\pi_{\Operators}}" left]
  \& \& {\IMode{\Arr{\Sets}}}
    \arrow[d, "\IMode{\pi_{cod}}"] \\
  {\IMode{\SCtx\times\Discrete{\Arities}}} \arrow[rr, "\IMode{\Operators}" below] \&
  \& {\IMode{\Sets}} \\
\end{tikzcd}
\]

@@

Next we have the interpretation of metavariable applications
$\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash\MApp{\MV{m}}{\vec{u}}{\vec{M}}
: \tau}_P$ where
$\Lookup{\Omega}{\MV{m}}{\MkValence{\vec{\sigma}}{\vec{\tau}}{\tau}}$, which
proceeds by projecting the metavariable's interpretation from the environment
and intantiating it via substitution:
\[
  \begin{tikzcd}[ampersand replacement = \&, sep=7em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
          \arrow[r, "\IMode{\Tuple{\pi_\MV{m}\,\pi_1, \phi\,\pi_2, \psi}}"]
    \& {\IMode{P_{\tau}^{\Yoneda{\vec{\sigma}\parallel\vec{\tau}}}\times S^{\{\vec{\sigma}\}} \times P^{[\vec{\tau}]}}}
          \arrow[r, "\IMode{\varsigma^\tau_{\vec{\sigma}\parallel\vec{\tau}}}"]
    \& \IMode{P_\tau}
  \end{tikzcd}
  \tag{metavar}
\]
where
\begin{gather*}
  \begin{tikzcd}[ampersand replacement = \&, sep=8em]
    {\IMode{S^{\{\Upsilon\}}}} \arrow[r, "\Define{\phi\,}{\,\Tuple{\pi_u}_{u\in\vec{u}}}"]
    \& {\IMode{S^{\{\vec{\sigma}\}}}}
  \end{tikzcd}\\
  \begin{tikzcd}[ampersand replacement = \&, sep=14em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
      \arrow[r, "\Define{\psi\,}{\,\Tuple{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash M:\tau}_P}_{(M,\tau)\in (\vec{M},\vec{\tau})}}"]
    \& {\IMode{P^{[\vec{\tau}]}}}
  \end{tikzcd}
\end{gather*}

The interpretation of operator applications is most complicated. Recall that,
unlike in standard treatments of universal algebra, our operators are indexed
by symbol collections; therefore, operators must pass through suitable
renamings in order to be used in the interpretation. Let us begin by
constructing for each operator $\IsOperator{\Upsilon}{\vartheta}{a}$ the
morphism
$\Of{\SemBrackets{\vartheta}_P}{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P
\to\Operators_{a}}$, which shall rename the parameters of the operator using
the environment. The natural transformation $\SemBrackets{\vartheta}_P$ is the
following composite:
\[
  \begin{tikzcd}[ampersand replacement = \&, sep=large]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}} \arrow[r, "\IMode{\pi_2}"]
    \& {\IMode{S^{\{\Upsilon\}}}} \arrow[r, "\IMode{\FLift{\pi_{(-)}}{}}"]
    \& {\IMode{\Operators_a}}
  \end{tikzcd}
\]

We will proceed using the initial $\SigFun{\Sigma}$-algebra $\alpha$, as
follows, by postcomposing it with a morphism $\beta$ from the environment into
the signature endofunctor, which interprets the syntax of operator
applications:
\[
   \begin{tikzcd}[ampersand replacement = \&, sep=large]
     {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}} \arrow[r, "\IMode{\beta}"]
     \& {\IMode{\SigFun{\Sigma}(P)_\tau}} \arrow[r, "\IMode{\alpha_\tau}"]
     \& {\IMode{P_\tau}}
   \end{tikzcd}
\]

The construction of $\beta$ proceeds by renaming the parameters of the operator
$\vartheta$ and constructing the (bound) arguments $\gamma$ of the operator.

\[
  \begin{tikzcd}[ampersand replacement = \&, sep=10em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
       \arrow[r, "\Define{\beta\,}{\,\Tuple{\SemBrackets{\vartheta}_P, \gamma}}"]
    \& {\IMode{%
       \coprod_{\vartheta\in \Operators_{\MkArity{\vec{v}}{\tau}}}%
       \prod_{\MkValence{\vec{\sigma}_i}{\vec{\tau}_i}{\tau_i}\in\vec{v}}%
       P_{\tau_i}^{\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}}%
    }}
  \end{tikzcd}
\]

Each argument
%
\begin{tikzcd}[ampersand replacement = \&]
  {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
    \arrow[r, "\IMode{\gamma_i}"]
  \& {\IMode{%
    P_{\tau_i}^{\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}}
  }}
\end{tikzcd}
%
is the exponential transpose of the composite,
\[
  \begin{tikzcd}[ampersand replacement = \&, sep=8em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P
       \times\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_u}
    }}
      \arrow[r, "\IMode{\Tuple{\pi_1\circ\pi_1,\phi_i,\psi_i}}"]
    \& {\IMode{\SemBrackets{\Omega\triangleright\Upsilon,\vec{\sigma}_i\parallel\Gamma,\vec{\tau}_i}_P}}
      \arrow[r, "\IMode{\SemBrackets{\Omega\triangleright\Upsilon,\vec{\sigma}_i\parallel\Gamma,\vec{\tau}_i\vdash M_i:\tau_i}_P}"]
    \& {\IMode{P_{\tau_i}}}
  \end{tikzcd}
\]
where $\phi_i,\psi_i$ are defined as follows:
\begin{gather*}
  \begin{tikzcd}[ampersand replacement = \&, sep=14em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P
       \times\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}
    }}
      \arrow[r, "\Define{\phi_i\,}{\,\Tuple{\pi_u\circ\pi_1}_{u\in\Dom{\Upsilon}},\, \Tuple{\pi_u\circ\pi_1\circ\pi_2}_{u\in\Dom{\vec{\sigma}_i}}}"]
    \& {\IMode{S^{\{\Upsilon,\vec{\sigma}_i\}}}}
  \end{tikzcd}\\
  \begin{tikzcd}[ampersand replacement = \&, sep=14em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P
       \times\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}
    }}
      \arrow[r, "\Define{\psi_i\,}{\,\Tuple{\pi_x\circ\pi_1}_{x\in\Dom{\Gamma}},\, \Tuple{\pi_x\circ\pi_1\circ\pi_2}_{x\in\Dom{\vec{\tau}_i}}}"]
    \& {\IMode{V^{[\Gamma,\vec{\tau}_i]}}}
  \end{tikzcd}
\end{gather*}

This concludes the interpretation of well-sorted terms into any $\Sigma$-model.
